\section{Compile-time verification: Pow with contracts}
\label{s:verification}
In this section we show how conventional verification techniques for OO can be applied on top of IC.
While IC is already presented in~\cite{servetto2014meta},
the idea and the techniques to use IC for verification are novel contributions.

Looking back at the code to generate \Q@pow@ in QQ and in IC,
we can
note how the method \Q@inductive(code)@ in IC is equivalent to the method
\Q@inductive(code)@ in QQ.
However,
the method \Q@inductive()@ in IC has no equivalent in QQ. The method
\Q@inductive()@ returns the complete code of a class with an abstract method,
that is then used as a handler to inject behaviour.
There is no way in traditional QQ to express this same concept,
and this is the crucial point that makes our approach easier to verify.
QQ is based on \emph{quasi} quotations, that is, \emph{parametric code} that
will become \emph{complete code} when you fill in the holes.
While in IC, every code literal is \emph{complete code} (with
the usual OO semantics where methods can be overridden).

We struggled to find literature about proving correct behaviour of parametric code (quasi quotations with holes). On the opposite, static verification of object oriented code,
is a well developed research area, where proving properties
of abstract classes is well-known.

We show this in the next example, where we handle \Q@pow@ as before, but while verifying that the
result encodes the right \Q@pow@ function. We will use similar notation to JML~\cite{jml}, with \Q!@ensure! and \Q!@result! representing methods pre and post conditions in special comments that are used by a verifying compiler and kept as metadata instead of ignored.
We include \Q!@ensureRV! for conditions that we expect to be verified at run time, instead of statically verified.
Many static verification approaches use some automatic theorem proving to figure out the proofs. Here, for completeness, we introduce a 
\Q! @proof! notation, where we insert the full proof, to show what we expect a theorem proving to generate.
To make the verification easier, we generate both the \Q@pow(x)@ method
and a \Q@exp()@ method, keeping track of the accumulated exponent.
This time we will use explicit iteration instead of recursion, just to show that our approach is not bound to any of those styles.
\newcommand\thisExp{\ensuremath{{}^{\textbf{this.exp()}}}}
\newcommand\thisSuperExp{\ensuremath{{}^{\textbf{this.superExp()}}}}
\newcommand\oneThisSuperExp{\ensuremath{{}^{\textbf{1+this.superExp()}}}}
\newcommand\powerY{\ensuremath{{}^{\textbf{y}}}}

\begin{lstlisting}
Pow={
  static method Library base()
    ={
    $\Comment{@ensures exp()=0}$
    method Num exp()=0 
    
    $\Comment{@ensures pow(x)=x\thisExp}$
    method Num pow(Num x)= 1 
    }
  ...
  }
\end{lstlisting}


\noindent
We start by examining the base case of our induction: we annotated our 
code with trivial specifications.
A code verifier could statically verify this code 
once and for all when \Q@Pow@ is type-checked.
Then, let us look at the \Q@inductive()@ step:

\begin{lstlisting}
Pow={
  static method Library base() =...$\Comment{as before}$
  static method Library inductive() 
    ={
    $\Comment{@ensures @result=1+this.superExp()}$
    method Num exp()=1+this.superExp() 
      
    method Num superExp()$\Comment{no body: it is an abstract method}$
      
    $\Comment{@ensures @result=x\thisExp}$
    $\Comment{@proof:}$
    $\Comment{ @result->x*this.superPow(x)//left side}$
    $\Comment{  =x* x\thisSuperExp = x\oneThisSuperExp}$
    $\Comment{ x\thisSuperExp->x\oneThisSuperExp//right side}$
    method Num pow(Num x)= x*this.superPow(x)
      
    $\Comment{@ensures pow(x)=x\thisSuperExp}$
    method Num superPow(Num x)$\Comment{no body: it is an abstract method}$
    }
  ...
  }
\end{lstlisting}

Here the annotations are much heavier.
We show the full proof that a theorem proving may generate.
Note that we need to rely on the contract of \Q@superPow(x)@ (a concept completely hidden in the QQ approach).
Note how also this code, and its possibly non-trivial proof, can be verified once and for all when the code of \Q@Pow@ is compiled.

While the library literals above have contracts,
neither methods \Q@base()@ nor \Q@inductive()@ have any contract:
statically we only know they return a \Q@Library@.

\noindent
Here we put the pieces together:

\begin{lstlisting}
Pow = {...
  $\Comment{@requiresRV y>=0}$
  $\Comment{@ensuresRV @result.exp().ensures = (@result = y)}$
  $\Comment{and @result.pow(x).ensures = (@result = x\powerY)}$
  static method Library generate(Num y){
    var Library res=this.base()
    for(i in Range(y)){
      res=Override[exp()<-superExp(), pow()<-superPow()](
        res,
        this.inductive())
    }
    return res } }
\end{lstlisting}

\noindent
When the \Q@Override@ operation  runs, it will
ensure the resulting class is statically verified:
every \Q@Library@ value has statically verified contracts. Operations 
that generate \Q@Library@ values (as \Q@Override@) will also compose those contracts, while verifying such
contract compositions are sound.
In our example
\Q@Override@ will check that 
\Q@res.superPow(x).ensures  satisfyAtLeast inductive.superPow(x).ensures@
(while renaming in \Q@res@ \Q@pow(x)@ and \Q@exp()@ as \Q@superPow(x)@ and \Q@superExp()@).
In our case it is true since they would be syntactically identical.
However, in other cases, this may of course fail.
We think of those operations as compile time operations lifted at execution.
\Q@Override@ producing an error is  equivalent to
getting compilation error about invalid \Q@extends@.
Note how this is not a form of run-time verification, but just the expected semantics of those operators.
\Q@Override@ does not guarantee to produce correct code all the time, in the same way a parser does not guarantee to produce a correct AST for all possible strings: producing errors is part of \Q@Override@ expected behaviour.

Finally, the \Q@return@ step triggers the run-time verification, and
checks the contracts of \Q@res@.
Before the return, we know that \Q@res@ is statically verified,
 but we have no guarantee of what its contracts are saying.
Indeed we are aiming for a run-time verified \Q@Pow.generate(y)@.
We believe this strikes a fundamental balance and is analogous to what a
verifying compiler~\cite{hoare2003verifying} should do.

This separation of concerns in our verification proposal is a key simplification
in our model:
all library values are well typed and statically verified all the time,
but what those contracts entail can be verified only at run-time, and
that the (meta-)programmer can choose when to do it.



\noindent
To summarize:
\begin{itemize}
\item
At compile time, once and for all, we statically verify that all
method bodies respect the (static) contracts on the method headers.
This can be handled exactly as static verification is normally handled in OO languages.
\item
During code composition we match contracts (\emph{satisfyAtLeast}).
There could be many possible designs for contract matching.
Here we rely on the simplest possible matching, following those two rules:
any contract \emph{satisfyAtLeast} the empty contract
and two syntactically equivalent contracts \emph{satisfyAtLeast} each other.
It can be surprising that those rules are sufficient;
the key is that 
the (meta-)programmer can write contracts
on purpose to make them matchable.
\item
We guarantee all library literals used by the execution are statically verified; but
statically we do not know the details of their contracts.
\item
Run-time verification is a simple and effective way to close this last gap, so that the result is not only \emph{self-consistent} but also the expected one.
\end{itemize}


%\section{OLD Introduction}
%
%Lisp, MetaML and other languages support metaprogramming by quasi quotation (QQ).
%It is a low-level technique: programmers think in terms of the desired AST;
%QQ can generate arbitrary shaped ASTs, thus allowing generation of \textbf{arbitrary behaviour}.
%
%We consider another technique, that we call incremental composition (IC), that
%also allows to generate arbitrary behaviour.
%IC enforces a high level of abstraction and do not work directly on the AST.
%While the unit of composition in QQ is the single AST node, 
%the unit of composition in IC is a \emph{Library}:
%a class or interface body, containing methods and possibly nested classes.
%Libraries are self contained in the sense that they contains no free variable.
%This cut at the root all scope-extrusion related problems, and (as shown later) enforces local reasoning.
%
%IC has been already presented in other works [...];
%in those works, IC expressive power is show by examples, but is not compared with QQ; moreover the tone  suggested IC power was inferior to QQ.
%
%Here we show that \textbf{IC expressive power is the same of QQ}.
%
%Moreover, 
%code generated by QQ need to be manually proved correct (after is generated).
%IC can be a little more verbose, but the correctness is verified during composition,
%thus result is, quite literally, correct by construction.
%In former work IC correctness was considered only in the limited sense of typing:
%Starting from well typed code, if IC produces a result, it is well typed by construction.
%
%Here we speculated how to reuse conventional Object oriented verification techniques so that
%\textbf{IC construct verified code}.






%-------------------------------------------------------------------------------------
%-------------------------------------------------------------------------------------

\section{IC is as expressive as QQ}
\label{s:pattern1}
So far we have presented how using pre and post conditions benefits the safety of our approach 
with respect to QQ.
\textbf{Independently} from those benefits, 
we now aim to show that IC is as expressive as QQ, and possibly easier to use on the large scale.

To this aim,
we will show how \Q@Override@
and other composition operators can be 
combined in a general purpose way to synthesize
new operators.

Except for the algebra of composition operators, that is already presented in~\cite{servetto2014meta},
all the considerations, ideas and techniques presented in this section are novel contributions.


%\subsection*{Object-orientation (OO) points to declarative languages}
%Imperative programming asks you to write down the
%detailed computational steps your machine should perform.
%This allows the programmer to reason about their
%programs by ``emulating computer'' in their mind.
%As we all know too well, this tempting approach to understand 
% program behaviour does not scale.
%We can see OO
%as taking imperative programming towards a more
%declarative style:
%\subsubsection*{Subtyping}
% A dynamically dispatched method invocation
%gives us no certainty of the detailed computational steps
%that are going to be performed, and
%requires programmers to reason in terms of the
%abstract contract of the method: that is
%the desired properties of the result as function of the arguments.
%The programmers are still in control of the detailed behaviour,
%but such control is delegated to the programmer that
%instantiated such object.
%The receiver object takes up the role of the \emph{reasoning engine} that takes
%care of transforming the programmer request into a result.
%In many OO languages, this reasoning engine amounts to just following a pointer. However,
% as the Visitor Pattern and other  design patterns show us, this is sufficient to encode very interesting behaviour resolution.
%Moreover, some languages implement multiple dispatch~\cite{clifton2000multijava}, making this reasoning engine a lot more expressive.
%\subsubsection*{Subclassing}

\subsection*{An algebra of composition operators}

In many class based object oriented languages, a class can inherit the code
from one or more parent classes.
In order to lift this capability as a metaprogramming operation,
we define an algebra of code composition where the values
are \Q@Library@ literals and the operations are the composition operators.

\begin{itemize}
\item A \Q@Library@ is a code literal:
a pair of balanced curly brackets with methods and nested classes inside.
Thanks to nested classes, a code literal can contain a large portion of code
with cooperating classes, possibly encapsulating a whole library.
\item A \emph{composition operator}
is a functionality taking in input \Q@Library@s and producing a \Q@Library@
result, or a \emph{composition error}.
\end{itemize}

\noindent
In the following example, class \Q@C@ contains a single static 
method \Q@foo()@ returning a \Q@Library@ literal.
We can use class \Q@C@ and the \Q@Override@ composition operator to
attempt creating classes \Q@D1,D2,D3@.

\begin{lstlisting}
C={
  static method Library foo()={ method Num m()=2 }  
  }
D1=C.foo()
D2=Override[](C.foo(),{method Num n()=0 })
D3=Override[m()<-superM1(),superM2()](C.foo(),C.foo(),{
  method Num m()=3+this.superM1()+this.superM2()
  $\Comment{superM1 and superM2 used for super calls}$
  method Num superM1()
  method Num superM2()
  })
\end{lstlisting}

\noindent
By a process known as flattening~\cite{scharli2003traits}, we get the following results for \Q@D1,D2,D3@:

\begin{lstlisting}
D1={ method Num m()=2 }  
\end{lstlisting}

\noindent
\Q@C.foo()@ is \emph{executed at compile time}
and we use the result to initialize \Q@D1@.

\begin{lstlisting}
D2={ method Num m()=2  method Num n()=0 }  
\end{lstlisting}

\noindent
\Q@Override@ works like inheritance; in this case since there is no conflict between
the two code literals, the result is a library literal containing both:
%\begin{lstlisting}
%D3=//composition error:
%//two implementations for method m()  
%\end{lstlisting}
%Here we try to compose the same code literal twice.
%This fails since both (identical) sides implement method m.

\begin{lstlisting}
D3={method Num m()=3+2+2 } $\Comment{super calls may be inlined}$
\end{lstlisting}

We write \Q@Override[m()<-superM1(),superM2()](..)@ to support calling the multiple conflicting versions
for method \Q@m()@.
Here we try to compose the same code literal twice.
This may fail, since both (identical) sides implement method m.
However, \Q@Override@ treats the last library value in a special
 and preferential way:
if there are multiple conflicting implementations of the same method 
(\Q@m()@ in this case)
the last value can redefine such method and avert the conflict.

The \Q@Override@ operator we are proposing works as 
trait composition~\cite{scharli2003traits}:
we can compose multiple \Q@Library@ values in
an associative and commutative way;
except for the last parameter, which
enjoys preferential composition:
its methods can override methods in the other
libraries. This privilege is equivalent to the 
one \emph{glue code} enjoys in the original trait model 
~\cite{scharli2003traits}(section 3.3).
Many variations of this composition operator has been presented in the literature,
and an exhaustive understanding of its behaviour is not needed to
understand this paper.

In addition to \Q@Override@, we will use other composition operators.
We believe it is fair to consider these high level operations on code as \emph{declarative}:
 the programmer does not specify the details of the source code
involved, but only reasons at the method call level.
Concretely, in the former example this means that the implementation of method \Q@m()@
inside class \Q@C@ is not relevant: only the behaviour/contract of such method is important.





%******************************
\subsection*{The concept of class decorators}

Conventional meta-programming techniques, based on AST rewriting by quote/unquote syntax
are very flexible, but push the user to generate the behaviour at the `source code' level of detail.
This is similar to the mindset of imperative programming, asking you to write down the
detailed computational steps your machine should perform.
This allows the programmer to reason about their programs by `emulating computer' in their mind.
As we all know too well, this tempting approach to understand  program behaviour does not scale.
Instead here we will use meta-programming only as a way to decide what operators to apply. Meta-programming will not directly generate code, but delegate this responsibility to operators.

Our approach
induces a desirable loss/transfer of control,
similar to the one happening in conventional inheritance, where
the programmer does not know the full set of methods a class offers:
this depends on the methods offered by the base class,
and if in a future release the heir is enriched with
more methods, such methods will also be injected into
the subclass.
For example, in

\begin{lstlisting}[language=Java]
class Rounded extends Button { ... }
\end{lstlisting}
\noindent we declare \Q@Rounded@s to be buttons in a way that
is parametric on what \Q@Button@s concretely are.
The programmer of \Q@Rounded@ just has to specify little bits
of behaviour to personalize the kind of button.
The concrete code is then automatically derived
by composing it with the existing code of \Q@Button@.
You can see \Q@Button@s as
following the double role of both a class and a class generator/decorator.
You can think of this as giving some suggestions to \Q@Button@
on what code to generate when creating the class \Q@Rounded@.
It is a very limited declarative language aimed to code composition.

In Java,  the \emph{code composition reasoning engine} is part of the language.
In Java it is a very simple minded reasoning, but it gets much more expressive in other languages, like C++.

Since the behaviour of inheritance
is set in stone, \Q@Button@ is not very active in deciding
about \Q@Rounded@.
However, this point of view lets us interpret many approaches
(traits~\cite{scharli2003traits}, mixins~\cite{smaragdakis2000mixin}, generics~\cite{igarashi2001featherweight}, family polymorphism~\cite{ernst2001family})
as ways to enrich what \Q@Button@ can do to generate the
required \Q@Rounded@ from the hints provided by the programmer:
a way to enrich the domain specific language
that the programmer uses to instruct base classes into generating their heirs.

In this article we wish to temporarily forget that base classes can be used as class or as types, and
focus on their active role in the generation
of code. In this context we will call them \emph{Class Decorators}.
Compile-time meta-programming~\cite{sheard2002template} is a good way to give Decorators
a mind of their own, so that they can perform arbitrary
complicated steps while generating their heirs.



%*******************








\subsection*{Patterns for generating arbitrary behaviour}
The general idea is to define a new operator
using other operators.
%(like Override for code composition and Refactor.redirect for generics).
Operators can generate imperative code, but such code is not directly
specified, it is synthesized by inductive reasoning.
%Instead of a detailed explanation of the behaviour of the operators we are going to use, we think it is more engaging
%to dive into the following code example:
As an example of a new operator, we will
consider \Q@Stringable@;
it can generate an opportune \Q@toS()@/\Q@toString()@ method inside of code literals.
In our example, the resulting code is going to be the definition
of class \Q@A@.
Note the declarative feel of this operation: we just specify the desired properties of the result (having a \Q@toS()@ method), and the detailed implementation is  
 automatically derived by the shape of the library literal provided.

\begin{lstlisting}
A=Stringable <>< { .. }
\end{lstlisting}

\subsection*{Babel fish operator \Q@<><@}
Since the main goal of our approach is to generate code, we introduce a \emph{code generation operator} \Q@<><@, called Babel fish.
We consider a language with normal operator overriding, where binary operators are just another syntactic representation for (static or not) method calls with a single parameter.
Babel fish is a binary operator, where the receiver takes up the role
of a `decorator' and the parameter is a \Q@Library@. The operator produces another \Q@Library@.
Thus \Q@Stringable <>< { .. }@ is equivalent to \Q@Stringable.babelFish({ .. })@ 



%Here \Q@Stringable@
%can generate a \Q@toS()@/\Q@toString()@ method inside of the code literal \Q@{ .. }@,
%the resulting code is going to be the definition
%of class \Q@A@.

We choose \Q@Stringable@ since its behaviour is not obvious: it is required to examine \Q@{ .. }@ in order to 
propagate \Q@toS()@ to all the fields.
This is also an important example: other operations like \Q@equals(that)@, \Q@hashCode()@ and \Q@compare(that)@
also rely on propagating the operation over the fields.
That is, all of these operations can be implemented by
following the same programming pattern we are going to show.

From now on, we will see \Q@Override@ as a class decorator, and
use it with syntax \Q@Override(a,b,c)<><d@ to highlight that the last parameter enjoys preferential composition.

\subsection*{Objective code}
In order to have a clear goal,
let's imagine a tentative code we would like to obtain:

\begin{lstlisting}
Stringable<><{
  S name; Num age;
  method S sayHi() = "Hi, i'm "+this.name();
  }
\end{lstlisting}

\noindent should evaluate into

\begin{lstlisting}
{S name; Num age;
 method S sayHi() = "Hi, i'm "+this.name();
 method S toS() 
   = "["+this.name().toS()+", "+this.age().toS()+"]";
}
\end{lstlisting}

How to obtain this kind of result by using IC instead of QQ?
The main idea is to try to not think about the actual code but about
its behaviour, and how to decompose it into functions, and how to
put these functions together.

%We will now show how the code of \Q@Stringable@ may look like.


\subsection*{Top level operation}
We are going to show the implementation of \Q@Stringable@ in a Java like language where
we are going to repeat parameter names in the call site when this can improve readability.
The following is our entry point: we define a babelFish static method
returning a \Q@Library@.

\begin{lstlisting}
Stringable={
  static method Library <><(Library that){
    libs=this.baseCases(that); $\Comment{step 1}$
    acc=this.fold(libs,acc:{method S toS()= ""}); $\Comment{step 2}$
    res=this.close(acc); $\Comment{step 3}$
    return Override[](res)<><that; $\Comment{step 4}$
  }
...
\end{lstlisting}


% The method is called \Q@<><@ and thus can be invoked as binary operator.
% We use the name \Q@Library@  for our code literals since they can contain nested classes, thus they potentially encapsulate a whole library. 

\noindent
The method uses inductive reasoning and is divided into 4 meaningful steps:
\begin{itemize}
\item[1] Using the input, we generate the base cases;
in this case computing \Q@toS()@ for a class with a \textbf{single specific field}.
That is, \Q@libs@ is a list, containing 
as many \Q@Library@ literals as the fields of \Q@that@.
Each literal has a \Q@toS()@ method that only handles a specific field.
\item[2] Then we  \textbf{fold} our base cases into a single solution; in this case we specify how to merge two \Q@toS()@ implementations.
\item[3] Finally we \textbf{close/wrap} our implementation (adding  \Q@[]@).
\item[4]
Our result \Q@res@ would now expose \textbf{only} the \Q@toS()@ method.
%(plus needed abstract methods and some private utility methods).
We then use \Q@Override@ to compose our \Q@toS()@ with the original input.
\end{itemize}

%We will see more of the details of \Q@Override@ while examing the rest of the code.

\noindent
These four steps are further explained below.

\subsection*{Step 1: base cases}
To define our base case, we first 
 declare a method returning a constant \Q@Library@ value: an approximation for a class
 whose  \Q@toS()@ method delegates to a field.

\begin{lstlisting}
static method Library baseTrait() = {
  T f;
  T = {method S toS();} $\Comment{nested class}$
  method S toS() = this.f().toS();
}
\end{lstlisting}

It has a field called \Q@f@ of type \Q@T@.
\Q@T@ is a type with an abstract \Q@toS()@ method returning a string \Q@S@.
The \Q@toS()@ method of the top level class, just propagates out
the behaviour of \Q@T.toS()@.

Note how we call such method a ``trait'': inspired by trait composition, 
this method represents a reusable piece of code, where all the dependencies are explicit.
That is, this code is \emph{general} (\Q@T@ offers only the required \Q@toS()@ signature)
but is not \emph{generic} in the Java/C\# sense.

\noindent
Now we can define our \Q@baseCases(that)@ method.

\begin{lstlisting}
static method Libs baseCases(Library that){
  res=Libs();//empty list
  for(fi in Fields(that)){
    tmp=RedirectType(path:"T" into:fi.type())<><
      Rename(selector:"f" into:fi.selector())<><
      this.baseTrait();
    res.add(tmp);
  }
  return res;
}
\end{lstlisting}

\noindent
We extract the fields by observing with introspection
our input class.
%As available in both Haskell and Python, we assume we can use the common syntax for list comprehension:
%\Q@CollectionType[e| x in list]@
%Where the expression \Q@e@ denotes the entries in the newly generated collection.
In this case we generate our entries with the following interesting code:

\begin{lstlisting}
RedirectType(path:"T" into:fi.type())<><
Rename(selector:"f" into:fi.selector())<><
this.baseTrait()
\end{lstlisting}

\noindent Here we apply two decorators to our \Q@baseTrait@. First
we rename our field name from \Q@f@ to the current field name, then
 we redirect \Q@T@ into the current field type.
 \Q@RedirectType@ is not an obvious operation, and is the kind of code manipulation where IC shines
with respect to QQ, since it handles logical problems that would end up hidden in QQ.
 The operator \Q@RedirectType@ is more general than \Q@Redirect@ as shown in~\cite{servetto2014meta}.
 \Q@Redirect@ is a powerful operator:
 it deletes the nested class \Q@T@ and replaces all
 the occurrences of \Q@T@ with occurrences of such external type.
 \Q@Redirect@ is indeed a powerful form of generics,
 where generic types can be expressed as nested classes with abstract members.
 However, logically \Q@Redirect(path:"T" into:fi.type())@
 would work only in case of \Q@fi.type()@ being a type
 defined \textbf{outside} of the decorated class.
 In case of \Q@fi.type()@ denoting a \textbf{nested class} inside of the decorated class,
 we would need to perform \Q@Rename(path:"T" into:fi.type())@ instead.
 That is,
 \Q@RedirectType@ can be expressed in the language itself as a \textbf{derived operator}
 that just switches between the two options and calls the right operation internally.
Exactly as  for \Q@Stringable@, once and for all the programmer can understand
 the complexity of a specific problem and encapsulate it in a solution.\footnote{
One may wonder what could the return type of \Q@fi.type()@
be. \Q@Rename@ and \Q@Redirect@ leverage on the
conventional nominal type system to avoid errors:
 \Q@fi.type()@ returns a general \Q@Type@ type, and offers operations to extract either
 the internal path (typed \Q@Path@, that would be a well typed argument for \Q@Rename@)
 or the external class (typed \Q@Class@, that would be a well typed argument for \Q@Redirect@).}
 On the other hand, while coding this kind of program manipulation with QQ, most programmers
 will just forget to handle the case where the type of a field is a nested class of the input.
 Depending on the target language, this could be a big problem later on.


\noindent
To give an example of the result of \Q@baseCases(that)@, if we were to call 

\begin{lstlisting}
baseCases({
  S name; Num age;
  method S sayHi() = "Hi, i'm "+this.name();
})
\end{lstlisting}

\noindent
we would obtain

\begin{lstlisting}
Libs[
  {S name;   method S toS() = this.name().toS();};
  {Num age;  method S toS() = this.age().toS();};
  ]
\end{lstlisting}

\noindent
Note how there is no trace of the \Q@sayHi()@ method, and the
nested classes named \Q@T@ have been removed to point at the
external types \Q@S@ and \Q@Num@.


\subsection*{Step 2: folding}
Next we fold all our base cases into a single \Q@toS()@ method.
Traditionally, to fold a list of values into a single value, a binary operation is needed.
However, here we fold code representing methods,
so we need a lifted version of fold:
 we supply a \Q@Library@ value where a method
uses two alternative versions of itself. 

\begin{lstlisting}
{
  method S toS()= this.superToS1()+", "+this.superToS2();
  method S superToS1();
  method S superToS2();
}
\end{lstlisting}

This corresponds both to multiple inheritance, where a new version of a method is obtained by composing the two super implementations,
but also to a binary fold operation from \Q@S superToS1()@ and \Q@S superToS2()@
into a \Q@S toS()@ result.
Thus, to apply this folding, we use \Q@Override@ and we leverage on the multiple inheritance interpretation:

\begin{lstlisting}
static method Library fold(Libs that,Library acc) {
  if (that.isEmpty()) {return acc;}
  newAcc=Override[toS()<-superToS1(),superToS2()](
      that.firstElement(),
      acc)<><{
      method S toS()= this.superToS1()+", "+this.superToS2();
      method S superToS1();
      method S superToS2();
    };
  return this.fold(that.withoutFirstElement(),acc:newAcc);
}
\end{lstlisting}

The base case of this recursive code is the empty list, where \Q@acc@ is returned,
otherwise \Q@Override@ provides us with multiple inheritance where
\Q@superToS1()@ and \Q@superToS2()@
allow us to call super from the first/second parameter.
Of course we can not just compose \Q@acc@ with the top of our list: they both offer
a \Q@toS()@ method. We need to provide extra code to override the conflicting implementation
and provide new behaviour, in this case, the two strings separated by a comma.

\noindent
Continuing with the example from before, starting from the same code literal we would now obtain:

\begin{lstlisting}
{S name; Num age;
 method S toS() = this.name().toS()+", "+this.age().toS();
}
\end{lstlisting}

\subsection*{Step 3: wrapping}
We could be satisfied with such result, but often we wish to wrap our final result
to present it better.

\begin{lstlisting}
static method Library close(Library that) =
  Override[toS()<-superToS()](that)<><{
    method S toS() = "["+this.superToS()+"]";
    method S superToS();
  }
}
\end{lstlisting}

\noindent
  In the \Q@close(that)@ method here, \Q@Override@ adds \Q@[]@ around the string.

\noindent
Starting from

\begin{lstlisting}
{S name; Num age; method S sayHi()= "Hi, I'm "+this.name();}
\end{lstlisting}

\noindent
we would now obtain

\begin{lstlisting}
{S name; Num age;
 method S toS()="["+this.name().toS()+", "+this.age().toS()+"]";
}
\end{lstlisting}

\subsection*{Steo 4: composition}
The last operation in the top level method, is composing our generated \Q@toS()@ behaviour with the
original code (containing also \Q@sayHi()@) in order to obtain the final result.

We wish to stress this example code could easily be adapted for \Q@equals(that)@ and
a plethora of other field dependent operations.
With slightly more adaptation it could generate any pattern based on method shape/names in
any code source.
This is clearly supporting all the expression power of MorphJ~\cite{huang2008expressive}.

It could be possible to ``abstract'' over this code, using the template method pattern so that
to write generators for, let's say, \Q@equals(that)@ and \Q@toS()@ the programmer may reuse the logic of \Q@fold(that,acc)@ and the top level method.
Here we preferred to show the concrete code of the Decorator \Q@Stringable@ for the sake of a more
direct example.

To summarize, in this section, we have a shown how a combination of
trait multiple inheritance, redirect/generics and rename
can be used as a starting point to synthesize arbitrary behaviour for a
code literal using an inductive mindset.
Trait multiple inheritance, redirect and rename
are high level class composition and adaptation operators,
and the obtained decorator also has a declarative `feel'.




%\section{Discussion}
%
%The approach presented here is very general, and could be adapted to fit many different languages.
%Those adaptations may have great impact on the general safety on the approach.
%
%The programming language 42 is based on this approach, and couple it with
%a strong nominal type system, where all the literal recursively contained in any executable code, are required to be well typed.
%However, operators still run on .... meta safety ...
%
%Note how composition operators can always fail with an exception, that ....
%
%Software verification conceptually try to match software with specifications.
%Software verification rarely consider metaprogramming,
%and is traditionally divided in two broad category:
%\begin{itemize}
%\item
%Static verification (the code is guaranteed to always give the right answer)
%and 
%\item Run-time verification (the code is guaranteed to never give the wrong answer, however it may non provide an answer, either by non-termination or by termination with an error).
%\end{itemize}
%When metaprogramming is took under consideration, those two category are not sufficient to express all the reasonable verification options;
%In particular, there is the issue of how the specification for the generated code (meta-specification) is obtained/generated.
%Here in the following some more possibilities:
%\begin{itemize}
%\item Full Static verification:
%The code is statically verified, thus always produce 
%correct code, that is also statically verified and 
%follow the Full Static verification criteria when generating
%new code.
%The specification contains a way to generated the meta-specification.
%This is the most hard core strategy and possibly the hardest to realize.
%
%\item Verified Model driven development:
%Meta-Specification are input of the (unverified) code generation process,
%the generated code is then statically verified to respect those specifications.  
%If the output is incorrect, it may be very hard to understand ``why''.
%
%\item Post-mortem static verification:
%A non verified system produce code and its meta-specification.
%Those specifications are then used to statically verify the result.
%Meta-specifications may be incorrect, but is easier to test that correctness of specifications instead of correctness of actual code.
%This is, for example, the approach of Gallina tactics for the Coq Theorem proving.
%
%\item Compile-time verification:
%The process of generating code is run-time verified,
%but this process happens in a phase before the generated code is executed.
%Thus this phase may fail, and the error may be regarded as a compilation error.
%However, if this phase succeed, the generated code is statically verified.
%(against correct meta-specifications, since  produced by a run-time verified process)
%In this approach, usually the generated code is unable to perform further steps of code generations.
%This approach is very powerful: the first phase may be considered as a phase of ``input validation'', and after that, the correct result is ensured.
%
%\item Deep-Run-time verification:
%The code is run-time verified, the generated code is run-time verified and so on.
%This is probably the simpler strategy to implement on top of an existing Run-time verification system
%
%\item Gradual verification:
%Like Deep-Run-time verification, but some algorithms may be simple enough that static verification is performed.
%In particular, the first layer of code generation may be statically verified.
%While gradual verification do not need to use the meta-generation
%boundaries to switch from Static to Run time verification, it may be a very common boundary.
% 
%\end{itemize}
