\PassOptionsToPackage{svgnames}{xcolor}
\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{VPT 2019} % Name of the event you are submitting to
%\usepackage{breakurl}             % Not needed if you use pdflatex only.
\usepackage{underscore}           % Only needed if you use pdflatex.
\input{../Def/Def}
\lstset{
    numbers=left,
    stepnumber=1,
    showstringspaces=false,
    firstnumber=last
}

\providecommand*{\code}[1]{\Q`#1`}
\newcommand{\saveSpace}{\vspace{-3px}}
\newcommand{\loseSpace}{\vspace{1ex}}
\usepackage{verbatim}

\title{Iteratively Composing Statically Verified Traits}
\def\titlerunning{Iteratively Composing Statically Verified Traits}

%magic code from https://tex.stackexchange.com/questions/344794/centering-issues-with-multiple-authors-with-the-same-affiliation-eptcs-format
\RequirePackage{array}
\newenvironment{authors}[1]%
  {\begingroup
   \gdef\estyle{}%
   \renewcommand\institute[1]%
     {\\\multicolumn{#1}{@{}c@{}}{\scriptsize\begin{tabular}[t]{@{}>{\footnotesize}c@{}}##1\end{tabular}}}%
   \renewcommand\email[1]%
     {\gdef\estyle{\footnotesize\ttfamily}\\##1\gdef\estyle{}}
   \begin{tabular}[t]{@{}*{#1}{>{\estyle}c}@{}}
  }%
  {\end{tabular}%
   \endgroup
  }

\def\anauthor#1#2{%
	#1%
	\institute{}
	\email{#2}%
}
\def\vuw{\institute{School of Engineering and Computer Science\\%
			Victoria University of Wellington\\%
			Wellington, New Zealand}}
\author{
	\begin{authors}{4}
	Isaac Oscar Gariano & Marco Servetto & Alex Potanin & Hrshikesh Arora
	\vuw
	\email{isaac@ecs.vuw.ac.nz & marco.servetto@ecs.vuw.ac.nz & alex@ecs.vuw.ac.nz & arorahrsh@myvuw.ac.nz}
	\end{authors}
}

\def\authorrunning{Isaac O.\,G., M. Servetto, A. Potanin \& H. Arora}

% Allows one to write /hello/ instead of \Q@hello@.
% Use // to get a normal text slash
\chardef\Slash=`\/
\catcode\Slash=\active
\chardef\other=12 % char code for other characters
\def/#1/{%
	\ifx/#1/% #1 is empty
		\Slash% just print a slash
	\else%
		\lstinline/#1/%
	\fi%
}
\let\oldinput=\input
\def\input#1{\oldinput{\detokenize{#1}}} % Don't expand commands in input (needed since / is a command)
\newcommand{\sref}[1]{Section~\ref{s:#1}}


%\definecolor{blue}{HTML}{0000F0} %
%\definecolor{purple}{HTML}{700090}
%\definecolor{orange}{HTML}{F07000}
%\definecolor{teal}{HTML}{0090B0}
%\definecolor{brown}{HTML}{A00000}
%\definecolor{green}{HTML}{008000}
%\definecolor{pink}{HTML}{F000F0}

\makeatletter
\renewcommand*\idstyle{%
	\expandafter\id@style\the\lst@token\relax}
\def\id@style#1#2\relax{%
        \ifcat#1\relax\else
                \ifnum`#1=\uccode`#1%
                        \color{DarkGreen}%
                \fi%
        \fi%
}
\makeatother


\lstset{%
	language=Java, morekeywords={exists, forall, @requires, @ensures, result, rename, hide, with},
	tabsize=2,
	identifierstyle=\idstyle,
%	aboveskip=0pt
%	keywordstyle=\color{blue},
%	commentstyle=\color{green},
%	stringstyle=\color{brown}
%	literate=
%		{||}{{$\vee$}}1
%		{&&}{{$\wedge$}}1
%		{<=}{{$\leq$}}1
%		{>=}{{$\geq$}}1
%		{**}{{$^{**}$}}1
}
%citations
\newcommand{\REV}[3]{%
	\NoteColour{red}{#1\NoteText{\footnote{%
		\textcolor{red}{\textbf{REV#2{:} #3}}}}}}
	
\begin{document}
\maketitle
\begin{abstract}
Static verification relying on an automated theorem prover can be very slow and brittle: since static verification is undecidable correct code may not pass a particular static verifier.
In this work we use metaprogramming to generate
code that is correct by construction.
A theorem prover is used only to verify the initial traits: little code unit that can be used to compose bigger programs.
In this way, trait composition is guaranteed to
start from correct code and to always produce correct code.
We extends conventional traits with methods pre and post conditions.
Crucially, the checking performed by the traditional trait composition (/+/)  operator need to also check the compatibility of contracts. In this way, there is no need to re-verify the produced code.

We show how to apply our approach to the
iconic example of the power function, where metaprogramming can generate fast (and correct by construction) specialized versions when the exponent known in advance.
\end{abstract}

%\input{abstract}
%\input{introduction}
%\input{background}
\input{background/static-verification}
\input{exampleExplanation}
%\input{combining}
%\input{examples}
%\input{conclusion}

\catcode\Slash=12% turn of my awesome slash
\bibliographystyle{eptcs}
\bibliography{paper}
\clearpage
\appendix
\input{appendix}
\end{document}
