\section{Introduction}
\emph{Stringly typed code} is
code where the predefined strings values and the string type are used excessively,
even when using a specialized datatype would be advantageous.
This often happens because most data source for a program are strings:
most field in databases are strings, all the data received from the network is seen as a 
string, and virtually all input the user insert from the keyboard is a string.
It is then more convenient while programming, to just pass the values around while keeping their type.
Moreover, in many languages defining a new type is very time consuming.
For examples, lets consider a database storing telephone numbers of customers.
Since databases do not have a special type for telephone numbers, they will be stored as strings.
When those strings are read by, let say, a Java program, a greedy programmer may just store
them as \Q@String@.
Originally, the programmer believed the only important operation on telephone numbers
was to print them back on a guy.
After some time, he had the need of extracting the prefix: lets declare a function to do so.
But... where this function should be? it can not be a member of the String type!
How many other similar functionalities may we need?
Finally, he understand that he should have inserted a new type 
\Q@TelephoneNumber@, where some functionality allows
to instantiate them from the string representation; their
\Q@toString@ method should just return the
string representation, and fields should contain
the country calling code, the extensions and other meaningful parts of the number.
Now, when the data is fetched from the DB, he have to write
\Q@TelephoneNumber.of(..)@, and same for the tests.

prefix, instances of such type must be , that should be initialized by taking a string, and should  
So, he go back, he refactored all his code and 
He just declared a function to such aim.

All languages with a minimal level of adoption allows
the user to write numbers and strings directly, 
as in \Q@1@ or \Q@"hi!"@.
Those special literals connect to types in the environment.
In some sense, those


There is no such thing as '1'

Cool

other talk: use/reuse

time
probably 20/30 min

topic: 
little: no literal -> all pay, all equal, stringable types ftw 
opt: sugar 
moral: turtles all the way down




\begin{acks}

\end{acks}
