\section{Proof and Axioms}
\label{s:proof}
\lstset{morekeywords={fwd}}

As previously discussed, instead of providing a concrete set of type rules, we provide a set of properties that the type system needs to respect.
We will express such properties using type judgements of the form $\ty{e}{T}$. This judgement form allows an $l$ to be typed with different types based how it is used, e.g. we might have $\S;\G;\h.m\oR l \cR \vdash l : \Kw{mut}\ C$  and $\S;\G;l.m\oR \h \cR \nvdash l : \Kw{mut}\ C$, where $m$ is a \Q!mut! method taking a \Q!read! parameter. Importantly, we allow types to change during reduction (such as to model promotions), but do not allow the types inside methods to change when they are called (see the \thm{Method Consistency} assumption below). However, for this to work in practice, the type-system may need to keep track of more information than what is provided by an \E and $\S$. Such information could include which method an $e$ in the main expression came from, or what previous reductions have been executed. Such extra information is not needed for the correctness of our proof, but rather to extend the set of valid programs, however doing so would add more verbosity to our notations.

\subheading{Auxiliary Definitions}
To express are type system assumptions, we first need some auxiliary definitions. We define what it means for an $l$ to be \reach from an expression:\\
\indent $\reach(\s, e, l)$ iff $\exists l' \in e$ such that $l \in \rog(\s, l')$.

\noindent We now define what it means for an object to be immutable: it is in the \rog of an \Q!imm! reference, or a \reach \Q!imm! field:\\*
\indent $\immut(\s, e, l)$ iff $\exists \E, l'$ such that:
\begin{iitemize}
\item $e = \E[l']$, \tyr{l'}{\Kw{imm}\,\_}, and $l \in \rog(\s, l')$, or
\item $\reach(\s, e, l')$, $\Ss(l').f = \Kw{imm}\,\_$, and $l \in \rog(\s, \s[l'.f])$.
\end{iitemize}

\noindent We define the \mrog of an $l$ to be the $l'$s recheable from $l$ by traversing through only \Q!mut! and \Q!capsule! fields:\\
\indent $l' \in \mrog(\s, l)$ iff $\exists f$ such that:
\begin{iitemize}
	\item $\Ss(l).f \in \{\Kw{capsule}\,\_, \Kw{mut}\,\_\}$, and
	\item $l' = \s[l.f]$ or $l' \in \mrog(\s, \s[l.f])$
\end{iitemize}

\noindent Now we can define what it means for an $l$ to be \muty by an expression $e$: something in $l$ is reachable from a \Q!mut! reference in $e$, by passing through any number of \Q!mut! and \Q!capsule! fields:\\
\indent $\muty(\s, e, l)$ iff $\exists \E,l'$ such that:
\begin{iitemize}
	\item $e = \E[l']$, $\tyr{l'}{\Kw{mut}\,\_,}$ and
	\item $\exists l'' \in \rog(\s, l)$ such that $l' = l''$ or $l'' \in \mrog(\s, l')$
\end{iitemize}

\noindent Note our \thm{Capsule Consistency} assumption below says that every \Q!capsule! reference is also \Q!mut!.

\noindent Finally we model the \encap property of \Q!capsule! references:\\
\indent $\encap(\s, \E, l)$ iff $\forall l' \in \rog(\s, l)$, if $\muty(\s, \E[\_], l')$, then 
not $\reach(\s, \E[\_], l')$.
We use the notation $\E[\_]$ to trivially transform an \E into an $e$; we expect $\E[\_]$ to be well-typed and not introduce any $l$s into the hole.


\subheading{Axiomatic Type Properties}
Here we assume the usual \thm{Subject Reduction}: any sub-expression of an $\s|e$ obtained from a well-typed and well-formed initial $\s_0|e_0$, is also well-typed:%
\SS\begin{Assumption}[Subject Reduction]\rm
	If $\VS(\s, \E[e])$, then $\Ss; \emptyset; \E \vdash e : T$.\\
\end{Assumption}

As we do not have a concrete type system, we need to assume some properties about its derivations. First we require that  \Q!new! expressions only have field initialisers with the appropriate type, fields are only updated with expressions of the appropriate type, methods are only called on receivers with the appropriate modifier, method parameters have the appropriate type, and method calls are typed with the return type of the method:%
\SS\begin{Assumption}[Type Consistency]\rm\ 
\begin{ienumerate}
%if S; G; E |- e.f = e' : _ C, and C.f = T', then S; G; E[e.f = []] |- e' : T'
\item If $C.i = T_i\,\_$, then \ty[\Kw{new}\ C\oR e_1,\ldots,e_{i-1},\h,e_{i+1},\ldots,e_n\cR]{e_i}{T_i}.
\item If \ty[\h.f \equals e']{e}{\_\,C} and $C.f = T'\,f$, then \ty[e.f \equals \h]{e'}{T'}.
\item If \ty[\h.m\oR e_1,\ldots,e_n\cR]{e}{\_\,C} and
$C.m = \mdf\,\Kw{method}\,\T\,m\oR\T_1\,\x_1,\ldots,\T_n\x_n\cR\,\_$, then:
\begin{enumerate}
\item \ty[\h.m\oR e_1,\ldots,e_n\cR]{e}{\mdf\,C},
\item \ty[e.m\oR e_1,\ldots,e_{i-1},\h,e_{i+1},\ldots,e_n\cR]{e_i}{T_i}, and
\item \ty{e.m\oR e_1,\ldots,e_n\cR}{T}.
%\item hello
\end{enumerate}
\end{ienumerate}
\end{Assumption}%

\noindent We also assume that any expression inside a method body, can be typed with the same type modifiers as when it is expanded by our \textsc{mcall} rule:%
\SS\begin{Assumption}[Method Consistency]\rm\
 If $\VS(\s, \EV[l.m\oR \vs \cR])$ %and $\s|\EV[l.m\oR v_1,\ldots,v_n \cR] \rightarrow \s|$,
where:
\begin{iitemize}
\item $\Ss; \emptyset; \EV[\h.m\oR \vs \cR] \vdash l : \_\,C$, $C.m =\_\,\Kw{method}\,\_\,m\oR\Many{T\,x}\cR\,\E[e]$, and
\item $\E'$ = $\M{l}{\E}{l.\invariant}$ if $C.m$ is a capsule mutator, otherwise $\E' = \E$,
\end{iitemize}

\indent then $\emptyset; \Kw{this} : \mdf\,C,\Many{x : T}; \E \vdash e : \mdf\,\_$ iff $\Ss; \emptyset; \EV[\E'[\Kw{this}\coloneqq l,\Many{x\coloneqq v}]] \vdash e[\Kw{this}\coloneqq l,\Many{x\coloneqq v}] : \mdf\,\_$.
% and $C.m  =\mdf\,\Kw{method}\,\_\,m\,\oR\T_1\,\x_1\ldots\T_n\x_n\cR\,\E[e]$ and
\end{Assumption}

\noindent Now we define formal properties about our TMs, thus giving them meaning. First we require that an \immut object not also be \muty: i.e. an object reachable from an \Q!imm! reference/field cannot also be reached from a \Q!mut!/\Q!capsule! reference and through \Q!mut!/\Q!capsule! fields:%
\SS\begin{Assumption}[Imm Consistency]\rm\ \\
\indent If $\VS(\s, e)$ and $\immut(\s, e, l)$, then not $\muty(\s, e, l)$.

\noindent Note that this dose not prevent \emph{promotion} from a \Q!mut! to a \Q!capsule!: a reduction step may change the type of an $l$ from \Q!mut! to \Q!imm!, provided that in the new state, the above assumption holds.
\end{Assumption}

We require that if something was not \muty, that it remains that way; this prevents, for example, runtime promotions from \Q!read! to \Q!mut!, as well as field access that return a \Q!mut! from a receiver that was not \Q!mut!:%
\SS\begin{Assumption}[Mut Consistency]\rm\ \\
\indent If $\VS(\s, \EV[e])$ and $\muty(\s, e, l)$ and $\s|\EV[e] \rightarrow^{+} \s'|\EV[e']$, then not $\muty(\s, e', l)$.
\end{Assumption}

\noindent We require that a \Q!capsule! reference be \encap; and require that \Q!capsule! is a subtype of \Q!mut!:%
\SS\begin{Assumption}[Capsule Consistency]\rm\ 
\begin{ienumerate}
\item If \tyr{l}{\Kw{capsule}\,\_}, then $\encap(\s, \E, l)$.
\item If \ty{e}{\Kw{capsule}\,C}, then \ty{e}{\Kw{mut}\,C}.
\end{ienumerate}
\end{Assumption}%

\noindent We require that field updates only be performed on \Q!mut! receivers:%
\SS\begin{Assumption}[Mut Update]\rm If \ty{e.f \equals e'}{T}, then \ty[\h.f \equals e']{e}{\Kw{mut}\,\_}.
\end{Assumption}

\noindent We also require that field access only be typed as \Q!mut!, if their receiver is also \Q!mut!:%
\SS\begin{Assumption}[Mut Access]\rm If \ty{e.f}{\Kw{mut}\,\_}, then \ty[\h.f]{e}{\Kw{mut}\,\_}.
\end{Assumption}

\noindent Finally, we require that a \Q!read! variable or method result not be typeable as \Q!mut!; in conjunction with \thm{Mut Consistency}, \thm{Mut Update}, and \thm{Method Consistency}, this allows one to safely pass or return a \Q!read! without it being used to modify the object's \rog:%
\SS\begin{Assumption}[Read Consistency]\rm\ 
\begin{ienumerate}
\item If $\G(x) = \Kw{read}\,\_$, then $\nty{x}{\Kw{mut}\,\_}$.
\item If \ty[\h.m\oR \es\cR]{e}{\_\,C} and $C.m = \_\,\Kw{method}\,\Kw{read}\,C'\,\_$, then $\nty{e.m\oR \es \cR}{\Kw{mut}\,\_}$.
\end{ienumerate}
\noindent Note that \thm{Mut Consistency} prevents an access to a \Q!read! field from being typed as \Q!mut!
\end{Assumption}\SS

\subheading{Strong Exception Safety}
Finally we assume strong-exception safety: the memory preserved by each \Q@try@--\Q@catch@ execution is not \muty within the \Q!try!:%
\SS\begin{Assumption}[Strong Exception Safety]\rm\ \\
\indent If $\VS(\s', \E[\Kw{try}^{\s_0}\oC e_0\cC\ \Kw{catch}\ \oC e_1\cC])$, then $\forall l \in \dom(\s_0)$, not $\muty(\s, e_0, l)$;.
\end{Assumption}

We use SES to prove that locations preserved by \Q@try@ blocks are never monitored (this is important as it means that a \Q!catch! that catches a monitor failure, will not be able to see the responsible object):%
\SS\begin{Lemma}[Unmonitored Try]\rm
	If $\VS(\s, \E[\Kw{try}^\s\oC\E[\M{l}{\_}{\_}]\cC\,\_])$, then $l\notin\s_0$
\end{Lemma}\SS
\begin{proof}
The proof is by induction: after 0 reduction steps, $e$ cannot contain a monitor expression by the definition of \VS. If this property holds for $\VS(\s, e)$ but not for $\s'|e'$ with $\s|e\rightarrow \s'|e$, we must have applied the \textsc{update}, \textsc{mcall}, or \textsc{new} rules; since our well-formedness rules on method bodies prevent any other reduction step from introducing a monitor expression. If the rule was a \textsc{new} rule, $l$ will be fresh, so it could not have been in $\s_0$. If the rule was an \textsc{update}, by \thm{Mut Update}, $l$ must have been \Q!mut!, similarly \textsc{mcall} will only introduce a monitor over a call to a \Q!mut! method, so by \thm{Type Consistency}, $l$ was \Q!mut!; either way we have that $l$ was \muty, so by \thm{Strong Exception Safety}, we have that $l \notin \s_0$.
\end{proof}

\subheading{Determinism}
We can use our object capability discipline (described in Section \ref{s:formalism}) to prove that the \Q!invariant! method is deterministic and does not mutate existing memory:%
\SS\begin{Lemma}[Determinism]\rm If $\VS(\s, \EV[l.\invariant])$ and
\begin{iitemize}
\item $\s|\EV[l.\invariant] \rightarrow \s'|\EV[e'] \rightarrow^{+} \s''|\EV[e'']$,
\end{iitemize}

\indent then $\s'' = \s,\_$, $\s|\EV[l.\invariant] \Rightarrow^+ \s''|\EV[e'']$, and $\forall l' \in \dom(\s)$, not $\muty(\s'', e'', l)$.
\end{Lemma}\SS
\begin{proof}
The proof will proceed by induction.

\emph{Base case}: If $\s|\EV[l.\invariant] \rightarrow \s'|\EV[e]$, then the reduction was performed by \textsc{mcall}.
By our well-formedness rules, the \Q!invariant! method takes a \Q!read! \Q!this!, so by \thm{Method Consistency} and \thm{Read Consistency}, we have that $l$ is not \muty in $e'$. By our well-formedness rules on method bodies and by \textsc{mcall}, we have that no other $l$ was introduced in $e'$, thus nothing is $\muty$ in $e'$.

The only non-deterministic single reduction steps are for calls to \Q!mut! methods on a \Q!Cap!; however \Q!invariant! is a \!read! methods, so even if $l$ = $c$, we have $\s|\EV[l.\invariant] \Rightarrow \s'|\EV[e']$. In addition, since \textsc{mcall} does not mutate $\s'$ with have $\s' = \s$.

\emph{Inductive case}: Consider $\s|\EV[l.\invariant] \Rightarrow^+ \s'|\EV[e'] \rightarrow \s''|\EV[e'']$. We inductively assume that $\forall l' \in \dom(\s)$, not $\muty(\s', e', l)$; thus by \thm{Mut Consistency}, each such $l'$ is not \muty in $e'$. We also inductively assume that $\s' = \s,\_$, since nothing in $\s$ was \muty, by \thm{Mut Update}, our reduction can't have modified anything in \s, i.e. $\s'' = \s,\_$. Finally, if $c \in \dom(\s)$, then it cant by $\muty$ in $e$, thus by \thm{Type Consistency}, we can't have called a \Q!mut! method on it. By definition of \Q!Cap!, no new instances of $c$ can be created, so no \Q!mut! methods of \Q!Cap! can be called; since calling such a method is the only way to get a non-deterministic reduction, we have $\s'|\EV[e'] \Rightarrow \s''|\EV[e'']$.
\end{proof}

%Thanks to how our reduction rules are designed, especially \textsc{try error},
%@Progress will need to rely on @StrongExceptionSafety internally.



% To the best of our knowledge, only the type system of 42~\cite{ServettoEtAl13a,ServettoZucca15}
%  supports all these assumptions out of the box,
% while both Gordon~\cite{GordonEtAl12} and Pony~\cite{clebsch2015deny,clebsch2017orca} supports all except StrongExceptionSafety,
% however it should be trivial to modify them to support it:
% the \Q@try-catch@ rule could be modified to
% $\emptyset;\G\vdash\Kw{try}\ \oC e_0\cC\ \Kw{catch}\ \oC e_1\cC:\T$
% if\\* $\emptyset;
% \G,\{x:\Kw{read}\,C | x:\Kw{mut}\,C\,\in\G\}
% \vdash e_0:\T$ and $\emptyset;\G\vdash e_1:\T$,
% i.e. $e_0$ can be typed when seeing all externally defined mutable references as \Q@read@.



\subheading{Proof of Soundness}
It is hard to prove \thm{Soundness} directly,
so we first define a stronger property,
called \thm{Stronger Soundness}, which depends on \mony and \OK. 

An object is \mony if execution
is currently inside of a monitor for that object, and
the monitored expression $e_1$ does not
contain $l$ as a \emph{proper} sub-expression:

\indent $\mony(e,l)$ iff
$e=\EV[\M{l}{e_1}{e_2}]$ and either $e_1=l$ or $l \notin e_1$.%\loseSpace

\noindent A monitored object is associated with an expression that can not observe it, but may
reference its internal representation directly.
In this way, we can safely modify its representation before checking its invariant.
The idea is that at the start the object will be valid and $e_1$ will reference $l$;
but during reduction, $l$ will be used to
modify the object; only after that moment, the object may become invalid.

\thm{Stronger Soundness} says that starting from a well-typed and well-formed $\s_0|e_0$, and performing any number of reductions, every \reach object is either \valid or \mony:%
\SS\begin{theorem}[Stronger Soundness]\rm
If \VS(\s, e) then $\forall l$, if $\reach(\s, e, l)$ then $\valid(\s, l)$ or $\mony(e, l)$.
\end{theorem}

\begin{theorem}\rm \thm{Stronger Soundness} $\Rightarrow$ \thm{Soundness}
\end{theorem}\SS
\begin{proof}
\noindent Suppose $\VS(\s, e)$, and $e = \EV[r_l]$:
\begin{ienumerate}
	% by \thm{Stronger Soundness}, \l is \valid or \mony?
	\item If $\valid(\s,l)$, then \l is valid, exactly as required.
	\item If $l$ is not $\valid$, and $l$ is \reach, by \thm{Stronger Soundness}, $\mony(e,l)$, $e = \E[\M{l}{e_1}{e_2}]$, and either:
	\begin{itemize}
	 \item $\EV = \E[\M{l}{\E'}{e_2}]$, that is $r_l$ (which is not equal to $l$) was found inside of $e_1$, this contradicts the definition of \mony, or
	 \item $\EV = \E[\M{l}{e_1}{\E'}]$, and thus $r_l$ was found inside $e_2$,  by our reduction rules, all monitor expressions start with $e_2=l.\invariant$, thus the first execution step
	 of $e_2$ is \trusted. Further execution steps are also \trusted, since by well-formedness the body of invariant methods only use \Q@this@ (now replaced with \l) to read fields.
	\end{itemize}
\end{ienumerate}
Since $l$ is \reach, either $l$ is $\valid$ or $r_l$ is \trusted.
\end{proof}

\subheading{Capsule Field Soundness}
Now we define and prove important properties about our novel \Q!capsule! fields. We first start with a few core auxiliary definitions. We first define a notation to easily get the \Q!capsule! fields of an $l$:\\
\indent $f \in \cf(\s, l)$ iff $\Ss(l).f = \Kw{capsule}\,\_$.

\noindent An $l$ is \HNC if it is not reachable from its \Q!capsule! fields:\\
\indent $\HNC(\s, l)$ iff $\forall f \in \cf(\s, l)$, $l \notin \rog(\s, \s[l.f])$.

\noindent We say that an $l$ is \WE if none of its \Q!capsule! fields is \muty without passing through $l$:\\
\indent $\WE(\s, e, l)$ iff $\forall f \in \cf(\s, l)$, not $\muty(\s\setminus l, e, \s[l.f])$.

\noindent We say that an $l$ is \NCM if we aren't in a monitor for $l$ that must have been introduced by \textsc{mcall}, and we don't access any of it's \Q!capsule! fields as \Q!mut!:\\
\indent $\NCM(\s, e, l)$ iff $\forall \E$:
\begin{iitemize}
\item if $e = \E[\M{l}{e'}{\_}]$, then $e' = l$. and
\item if $e = \E[l.f]$, $f \in \cf(\s, l)$, and \ntyr[\h.f]{l}{\Kw{capsule}\,\_}, then \ntyr{l.f}{\Kw{mut}\,\_}.
\end{iitemize}

\noindent Finally we say that $l$ is \HNO if we are in a monitor introduced for a call to a capsule mutator, and $l$ is not reachable from inside this monitor, except perhaps through a single \Q!capsule! field access.\\
\indent $\HNO(\s, e, l)$ iff $e = \EV[\M{l}{e'}{\_}]$, and either:
\begin{iitemize}
\item $e' = \E[l.f]$, $f \in \cf(\s, l)$, and not $\reach(\s, \E[\_], l)$ or
\item not $\reach(\s, e', l)$.
\end{iitemize}

\noindent Now we formally state the core propties of our \Q!capsule! fields (informally described in \ref{s:protocol}):%
\SS\begin{theorem}[Capsule Field Soundnes]\rm
If $\VS(\s, e)$ then $\forall l$, if $\reach(\s, e, l)$, then:\\
\indent $\HNC(\s, l)$ and either:
\begin{iitemize}
\item $\WE(\s, e, l)$ and $\NCM(\s, e, l)$, or
\item $\HNO(\s, e, l)$.
\end{iitemize}
\end{theorem}\SS
\begin{proof}
This trivially holds in the base case when $\s = c\mapsto\Kw{Cap}\{\}$, since \Q!Cap! has no \Q!capsule! fields.
Now we suppose it holds for a $\VS$ and prove it for the next $\VS$.
 
Note that any single reduction step can be obtained by exactly one application of the \textsc{ctxv} rule and one other rule.
We will first proceed by cases on the property we need to prove, and then by the non-\textsc{ctxv} reduction rules that could violate or ensure it:%
\begin{ienumerate}
\item \HNC:
\begin{enumerate}
	\item (\textsc{new}) $\s|\EV[\Kw{new}\ C\oR v_1,\ldots,v_n\cR]\rightarrow \s'| \EV[\M{l}{l}{l\singleDot\invariant}]$, where $\s' = \s,l\mapsto C\{v_1,\ldots,v_n\}$:
	\begin{itemize}
		\item This reduction step doesn't modify any pre-existing $l'$, so we can't have broken \HNC for them.
		\item Since the pre-existing \s was not modified, by \VS, $l \notin \rog(\s, v_i) = \rog(\s', \s'[l.f])$; thus \HNC holds for $l$.
	\end{itemize}

	\item (\textsc{update}) $\s|\EV[l.f\equals{}v]\rightarrow \s[l.f=v]|\EV[\M{l}{l}{l.\invariant}]$:
	\begin{itemize}
		\item Consider any $l' \neq l$:
		\begin{itemize}
			\item If $l'$ was \WE, by \thm{Mut Update}, $l$ is \Q!mut!. Consider any $f' \in \cf(\s, l')$, by \WE, the \rog of $l'.f'$ is not \muty and so $l \notin \rog(\s, \s[l'.f'])$, so we can't have modified $\rog(\s, \s[l'.f'])$, hence $l$ is still \HNC.
			\item Otherwise, $l'$ was \HNO, and so $l' \notin \rog(\s, v)$, so we can't have added $l'$ to the \rog of any field, thus \HNC still holds.
		\end{itemize}
%		\item $l$ cant have been \HNO, since we had $l.f = v$ inside our evaluation context, and $l.f = v$ is not a field access.
		\item If $f \in \cf(\s, l)$:  by \thm{Mut Update}, we have that $l$ is \muty, so by \thm{Type Consistency} and \thm{Capsule Consistency}, $\encap(\s, \EV[l.f = \square], v)$, hence $l$ is not \reach from $v$, and so after the update, \HNC holds for $l$.
		\item By the proof above for the case $l' \neq l$, we can't have broken \HNC for $f \notin \cf(\s, l)$, nor any other field of $l$.
	\end{itemize}
	\item No other reduction rule modifies memory, so they trivially preserve \HNC for all $l$s.
\end{enumerate}

\item \HNO:
\begin{enumerate}
	\item (\textsc{access}) $\s|\EV[l.f]\rightarrow \s|\EV[\s[l.f]]$:
	\begin{itemize}
		\item Suppose $l$ was \HNO, then $\EV = \EV'[\M{l}{\E[l.f]}{\_}]$, with $l$ not \reach from $\E$, and $l.f$ is an access to a \Q!capsule! field. By \HNC, $l$ is not in the \rog of $\s[l.f]$, and so $l$ is not \reach from $\E[\s[l.f]]$, and so \HNO still holds.
		\item Clearly this reduction cannot have made any $l'$ \reach in a sub-expression where it wasn't allready \reach, so we can't have violated \HNO for any other $l'$.
	\end{itemize}


	\item (\textsc{monitor exit}) $\s|\EV[\M{l}{v}{\Kw{true}}]\rightarrow \s|\EV[v]$:
	\begin{itemize}
		\item As with the above case, we can't have violated \HNO for any $l' \neq l$.
		\item If this monitor was introduced by \textsc{new} or \textsc{update}, then $v = l$. And so \HNO can't have held for $l$ since $l = v$, and $v$ was not the receiver of a field access.
		\item Otherwise, this monitor was introduce by \textsc{mcall}, due to a call to a capsule mutator on $l$. Consider the state $\s_0|\EV[e_0]$ immediately before that \textsc{mcall}:
		\begin{itemize}
			\item We must not have had that $l$ was \HNO, since $e_0$ would contain $l$ as the receiver of a method call. Thus $l$ was originally \WE and \NCM.
			\item Because \NCM held in $s_0|\EV[e_0]$, and $v$ contains no field accesses or monitor, it also holds in $\EV[v]$.
			\item Since a capsule mutator cannot have any \Q!mut! parameters, by \thm{Type Consistency}, \thm{Mut Consistency}, and \thm{Mut Update}, the body of the method can't have modified $\s_0$: thus $\s = \s_0, \_$. Since no pre-existing memory has changed since the \textsc{mcall}, and a capsule mutator cannot have a \Q!mut! return type, by \thm{Type Consistency}, we must have $\Ss; \emptyset; \EV \vdash v : \mdf\,\_$ where $\mdf \neq \Kw{mut}$:
			\begin{itemize}
				\item If $\mdf = \Kw{capsule}$, by \thm{Capsule Consistency}, the value of any \Q!capsule! field of $l$ cant be in the \rog of $v$ (unless $l$ is no longer \reach), so we haven't made such a field \muty.
				\item Otherwise, $\mdf \in \{\Kw{read}, \Kw{imm}\}$, by \thm{Read Consistency}, \thm{Imm Consistency}, and \thm{Mut Consistency}, we have that $v$ is not \muty.
			\end{itemize}
			Either way, the \textsc{monitor exit} reduction has restored $\WE(\s_0, \EV[e_0], l)$.
		\end{itemize}
	\end{itemize}
	
	\item No other rules remove monitors, field accesses or make something \reach that wasn't before; thus they preserve \HNO for all $l$s.
\end{enumerate}

\item \NCM:
\begin{enumerate}
		\item (\textsc{mcall}) $\s|\EV[l\singleDot\m\oR v_1,\ldots,v_n\cR]\rightarrow \s|\EV[e[\Kw{this}\coloneqq l,\x_1\coloneqq v_1,\ldots,x_n\coloneqq v_n]]$:
		\begin{itemize}
			\item Suppose $m$ is not a capsule mutator, by our well-formedness rules for method bodies, $e$ doesn't contain a monitor.
			\begin{itemize}
			\item Since $m$ is not a capsule mutator, if $e = \E[l.f]$, for some $f \in \cf(\s, l)$, we must have that $m$ was not a \Q!mut! method. So by \thm{Mut Access} and \thm{Method Consistency}, we have that $\Ss; \emptyset; \EV[\E] \nvdash l.f : \Kw{mut}\,\_$ only if $m$ was a \Q!capsule! method, but by \thm{Method Consistency} that would mean that $\Ss; \emptyset; \EV[\E[\h.f]] : \Kw{capsule}\,\_$. Either way, we cant have broken \NCM for $l$.
			\item Consider $l' \neq l$, since fields are instance private, and by our well-formedness rules on method bodies, $l' \notin e$, we can't have introduced any field accesses $l$. As $e$ doesn't contain monitor either, we haven't broken \NCM for $l'$.
		\end{itemize}
		\item Otherwise, $e = \M{l}{e'}{l.\invariant}$. By our rules for capsule mutators, $m$ must be a \Q!mut! method with only \Q!imm! and \Q!capsule! parameters, thus by \thm{Type Consistency}, $l$ must have been \Q!mut!, and each $v_i$ must be \Q!imm! or \Q!capsule!. By \thm{Imm Consistency} and \thm{Capsule Consistency}, $l$ can't be reachable from any $v_i$. Since capsule mutators use \Q!this! only once, to access a \Q!capsule! field, $e' = \E[l.f]$, for some $f \in \cf(\s, l)$. Since $l$ is not \reach from any $v_i$, $l \notin \E$, by our well-formedness rules for method bodies, $l$ is not \reach from any $l' \in \E$, and so \HNO now holds for $l$.
		\end{itemize}	
		\item Since no other rule can introduce a monitor expression over an $e \neq l$, nor introduce field access, by \thm{Mut Consistency} and \thm{Mut Access}, we can't have broken \NCM for any $l$.
\end{enumerate}

\item \WE:
	\begin{enumerate}
		\item (\textsc{new}) $\s|\EV[\Kw{new}\ C\oR v_1,\ldots,v_n\cR]\rightarrow \s,l\mapsto C\{v_1,\ldots,v_n\}| \EV[\M{l}{l}{l\singleDot\invariant}]$:
		\begin{itemize}
			\item Consider any pre-existing $l'$. Suppose we broke \WE for $l'$ by making some $f' \in \cf(\s, l)$ \muty. Since the \rog of $l'$ can't have been modified, nor could the \rog of any other pre-existing $l''$, we must have that $\s[l'.f]$ is now \muty through some $l.f$. This requires that a $v_i$ be an initialiser for a \Q!mut! or \Q!capsule! field, which by \thm{Type Consistency} and \thm{Capsule Consistency}, means that $v_i$ must also be typeable as \Q!mut!. But then the $\s[l'.f]$ was already \muty through $v_i$, so $l'$ can't have already been \WE, a contradiction.

			\item Now consider each $i$ with $C.i = \Kw{capsule}\,\_\,f$. By \thm{Type Consistency} and \thm{Capsule Soundness}, $v_i$ was \encap and $\rog(s, v_i)$ is not \muty from \EV, and so $v_i$ is not $\muty(\s'\setminus l, \EV[\M{l}{l}{l.\invariant}], v_i)$; thus \WE holds for $l$ and each of its capsule fields.
		\end{itemize}
		
		
		\item (\textsc{update}) $\s|\EV[l.f\equals{}v]\rightarrow \s[l.f=v]|\EV[\M{l}{l}{l.\invariant}]$:
		\begin{itemize}
			\item Consider any $l' \neq l$, where $l'$ was \WE. Consider any $f' \in \cf(\s, l')$, by \WE, the \rog of $l'.f'$ was not \muty (except through $l'$), since by \thm{Mut Update}, $l$ is \Q!mut!, $l' \notin \rog(\s, \s[l'f])$, so we can't have modified $\rog(\s, \s[l'.f'])$; so $l'.f'$ can't now be \muty through $l$. By \thm{Mut Consitency}, we couldn't have have made $l'.f'$ \muty some other way, so $l'$ is still \WE.
			\item If $l$ was \WE, and $f \in \cf(\s, l)$. By \thm{Type Consistency} and \thm{Capsule Consistency}, $v$ is \encap, thus $v$ is not \muty from \EV, and $l$ is not \reach from $v$, thus $v$ is still \encap and \WE still holds for $l$ and $f$.
			\item By the proof above for the case $l' \neq l$, we can't have broken \WE for $f \notin \cf(\s, l)$, nor any other field of $l$.
		\end{itemize}
		
		
		\item (\textsc{access}) $\s|\EV[l.f]\rightarrow \s|\EV[\s[l.f]]$:
		\begin{itemize}
			\item Suppose $l$ was \WE and \NCM, and $f \in \cf(\s, l)$, by \thm{Mut Access}, either $\Ss; \emptyset; \EV \nvdash \s[l.f] : \Kw{mut}\,\_$ or $\Ss; \emptyset; \EV[\h.f] \vdash l : \Kw{capsule}\,\_$. If $l$ was \Q!capsule!, then by \thm{Capsule Consistency} and \HNC, $l$ is not \reach from $\EV[\s[l.f]]$, so it is irrelevant if $l$ is no longer \WE. Otherwise, if $l$ was not \Q!capsule!, $\s[l.f]$ will not be \Q!mut!, so \WE is preserved for l. Note that if $l$ wasn't \NCM, it was \HNO, so we don't need to preserve \WE. 
			\item Since this reduction doesn't modify memory, by \thm{Mut Consistency}, there is no other way to make the \rog of a capsule field $f'$ of $l'$ \muty, without going through $l'$, so \WE is preserved for $l'$.
		\end{itemize}
	\end{enumerate}
	
	\item Since none of the other reduction rules modify memory, by \thm{Mut Consistency}, they cant violate \WE.
	
\end{ienumerate}

\noindent In each case above, for each $l$, \HNC holds; and either \WE and \NCM holds, or \HNO holds.
\end{proof}

\subheading{Proof of Stronger Soundness}
\begin{proof}
We will prove this inductively, in similar way to how we proved \thm{Capsule Field Soundness}.
In the base case, we have $\s = c\mapsto\Kw{Cap}\{\}$, since \Q!Cap! is defined to have the trivial invariant, we have that $c$ (the only thin in \s), is \valid.

Now we assume that everything reachable from the previous \VS was was \valid or \mony, and proceed by cases on the non-\textsc{ctxv} rule that gets us to the next \VS.
\begin{ienumerate}

	\item (\textsc{update}) $\s|\EV[l.f\equals v]\rightarrow \s'|\EV[e']$, where  $e'=\M{l}{l}{l.\invariant}$:
	\begin{itemize}
		\item Clearly $l$ is now $\mony$.
		\item Consider any other $l'$, where $l \in \rog(\s,l')$ and $l'$ was \valid; now suppose we just made $l'$ not \valid. By our well-formedness criteria, \Q@invariant@ can only accesses \Q@imm@ and \Q@capsule@ fields, thus by \thm{Imm Consistency} and \thm{Mut Update}, we must have that $l$ was in the \rog of $l'.f'$, for some $f' \in \cf(\s, l')$. Since $l \neq l'$, $l'$ can't have been \WE. Thus, by \thm{Capsule Field Soundness}, $l'$ was \HNO, and $\EV = \EV'[\M{l'}{\EV''}{\_}]$:
		\begin{itemize}
			\item If $\EV''[l.f \equals v] = \E[l'.f']$, then by \HNO, $l'$ is not reachable from \E. The monitor must have been introduced by an \textsc{mcall}, on a capsule mutator for $l'$. Since a capsule mutator can take only \Q!imm! and \Q!capsule! parameters, by \thm{Type Consistency}, \thm{Imm Consistency}, and \thm{Capsule Consistency}, $l$ cannot be in their \rog{}s (since $l$ was in the \rog of $l'$, and $l$ is \Q!mut!). Thus the only way for the body of the monitor to access $l$ is by accessing $l'.f'$.
			Since capsule mutators can access \Q!this! only once, and by the proof of \thm{Capsule Field Soundness}, there is no other $l'.f'$ in $\E[l'.f']$, nor was there one in a previous stage of reduction, hence $l$ is not \reach from $\E$. This contradicts the that we just updated $l$.
			\item Thus, by \HNO, we must have $\EV''[l.f \equals v] = e$, with $l'$ not \reach from $e$; so $l'$ was, and still is, \mony.
		\end{itemize}
		\item Since we don't remove any monitors, we can't have violated \mony. In addition, if a $l$ was not reachable from a \valid $l'$, it is still \valid.
	\end{itemize}

	\item (\textsc{access}) $\s|\EV[l\singleDot f] \rightarrow \s|\EV[v]$:
		\begin{itemize}
			\item Consider any \mony $l'$, then $\EV = \EV'[\M{l'}{\EV''}{l.\invariant}]$. The monitor can't have been introduced by \textsc{new} or \textsc{update}, since they wrap monitors around $v$s which cannot be reduced; thus the monitor was introduced by \textsc{mcall}.
				Since the monitor for $l'$ contains $l.f$, it cant have been \NCM, thus by \thm{Capsule Field Soundness}, we must have that $l'$ is \HNO. 
				Thus $\s[l.f] \neq l'$, and so $l'$ is still \mony.
			\item Otherwise $l'$ was \valid, since this rule doesn't mutate memory, by \thm{Determinism}, it is still \valid.
		\end{itemize}

	\item (\textsc{monitor exit}) $\s|\M{l}{v}{\Kw{true}}\rightarrow \s|v$: By our \VS and our well-formedness requirements on method bodies, the monitor expression must have been introduced by \textsc{update}, \textsc{mcall}, or \textsc{new}. In each case the 3\textsuperscript{rd} expression started of as $l.\invariant$, and it has now (eventually) been reduced to $\Kw{true}$, thus by \thm{Determinism} $l$ is \valid.
	
		\item (\textsc{new}) $\s|\EV[\Kw{new}\ C\oR\vs\cR]\rightarrow \s,l\mapsto C\{\vs\}|\EV[ \M{l}{l}{l.\invariant}]$:
			Clearly the newly created object, $l$, is \mony. This rule does not modify pre-existing memory, introduce pre-existing $l$s into the main expression, nor remove monitors, thus every pre-existing $l'$ is still \valid (due to \thm{Determinism}), or \mony.
			
		\item (\textsc{mcall}, \textsc{try enter}, \textsc{try ok}, and \textsc{try error}):
			These reduction steps do not modify memory, the memory locations reachable inside of the main expression, or any pre-existing monitor expressions. Thus by the case for \textsc{new} above, they could not have violated \valid or \mony for any $l$.
\end{ienumerate}

\noindent Regardless of the reduction performed, we have each \reach $l$ is \valid or \mony.
\end{proof}